---
title: "HW_2"
author: "Nathan Laroy"
date: "2023-11-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0) prepare data

```{r}
NB <- c(15.4, 13.5, 13.3, 12.4, 12.8, 13.5, 14.5, 13.9,
        11.0, 15.0, 17.0, 13.8, 17.4, 16.5, 14.4)

EFB<- c(16.5, 13.2, 13.6, 13.6, 14.0, 14.0, 16.0, 14.1,
        11.5, 14.4, 16.0, 13.2, 16.6, 18.5, 14.5)

CBF_data <- data.frame("Subject" = c(1:15),
                       "NB" = NB,
                       "EFB" = EFB)

```
# 1) graphing and exploratory data analysis
#### GRAPHS
```{r}
# standard scatterplot with LS line
plot(CBF_data$NB, CBF_data$EFB)
abline(lm(CBF_data$EFB ~ CBF_data$NB))

# scatterplot of differences with smoother (ideally about zero)
diff <- CBF_data$EFB - CBF_data$NB
plot(diff)
lines(x = 1:15, y = rep(0, times=15))
lines(lowess(diff), col = "red")

# boxplot of differences with horizontal about mean
boxplot(diff)
abline(h = mean(diff), col = "red", lwd = 3)

#### EXPLORATORY DATA ANALYSIS
# central Q: explore data wrt association
mean(diff)
sd(diff)

# other: in reference to graphs
```

# 2) Estimate Spearman Rank Correlation
The Spearman rank correlation is estimated at: `r cor(CBF_data$EFB, CBF_data$NB, method = "spearman")`.

The Spearman rank correlation coefficient is calculated as follows: first, both variables are ordered from least to greatest and ranked accordingly. Potential ties for each batch of observations are resolved via mid-ranks (i.e., the 'normal' ranks of the ties are averaged across the number of constituents of the respective tie). The coefficient is then mathematically defined as:
$$
\begin{aligned}
r_s = \frac{12\sum_{i=1}^{n}\left(\left[R_i - \frac{n + 1}{2}\right]\left[S_i - \frac{n + 1}{2}\right]\right)}{n(n^2-1)}
\end{aligned}
$$
with $R_i$ the ranks of the first variable's observations, $S_i$ the ranks of the second variable's observations, and $n$ the number of paired observations.

# 3) Permutation test based on Spearman's rank statistic

What does $H_0$ imply?

Under $H_0$, if there is no correlation between the NB and EFB measurements, then any observed value of NB can co-occur with any observed value of EFB.

To perform a permutation test, we first construct a permutation null distribution. What is to be permuted exactly? Given that under $H_0$ any NB value may co-occur with any EFB value, it is easiest to permute the NB vector, and calculate iteratively the Spearman correlation coefficient for each combination of EFB and permuted NB.

```{r}
set.seed(44)
obs.cor <- cor(NB, EFB, method = "spearman")

N <- 10000
spearman.r.star <- replicate(N, {
  NB_new <- sample(NB)
  spearman.r <- cor(NB_new, EFB, method = "spearman")
  return(spearman.r)
})
CritValues <- quantile(spearman.r.star, probs = c(0.95))
hist(spearman.r.star, breaks = 40)
abline(v = c(CritValues,obs.cor), col = c("red","blue")
                              , lwd = 2)
print(c(max(spearman.r.star), obs.cor))
sum(spearman.r.star >= obs.cor)
mean(spearman.r.star >= obs.cor)
```

# 4) Asymptotic approximation
## standardized coefficients (with ties!)
```{r}
n <- 15
NB_ranked <- rank(NB) # 1 tie: 5.5 (2)
EFB_ranked <- rank(EFB) # 4 ties:
#                              2.5 (2)
#                              4.5 (2)
#                              6.5 (2)
#                              11.5 (2)
Diff_ranked <- NB_ranked - EFB_ranked
g <- 1
t_i <- 2
h <- 4
u_i <- 2
## r.star == obs.cor, because R calculates tie-corrected by default
r.star <- (n*(n^2 - 1) - 6*sum(Diff_ranked^2) - 0.5*(g*(t_i*(t_i^2 - 1)) + 
                                                       h*(u_i*(u_i^2 - 1)))) / 
  (((n*(n^2 - 1) - g*(t_i*(t_i^2 - 1))) * (n*(n^2 - 1) 
                                           - h*(u_i*(u_i^2 - 1))))^0.5)

r.standard <- (n-1)^(0.5) * r.star
(p.value <- 1 - pnorm(r.standard))

# can be doublechecked via perm.cor.test from "jmuOutlier" package)
```

# 5) Compare two distributions
```{r}
probabilities <- seq(0.0001, 0.9999, by = 0.01)
probabilities[1:5]

perm.Quantiles <- quantile(spearman.r.star, probabilities)
n.Quantiles <- qnorm(probabilities)

plot(n.Quantiles, perm.Quantiles)
qqline(perm.Quantiles)

# or
qqnorm(perm.Quantiles)
qqline(perm.Quantiles)
```

